---
title: "MIRA 16S Preliminary Analysis (DADA2)"
author: "Erik Clarke"
date: "6/1/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  results="hide",
  warning=FALSE,
  message = FALSE,
  fig.width=5,
  fig.height = 3)

library(tidyverse)
library(dada2)

ROOT_DATA_FP <- "~/projects/M01_Shared_Data"
ROOT_OUT_FP <- "~/projects/M01-01_MIRA_16S_DADA2/dada2_output"
FIG_FP <- "~/projects/M01-01_MIRA_16S_DADA2/Figures"

mira_theme <- theme_bw() + theme(
  strip.background = element_rect(fill = NA)
)
theme_set(mira_theme)
```

```{r load-data}
load_all_data <- function() {
  seqtab <- readRDS(file.path(ROOT_OUT_FP, "seqtab.rds"))
  taxa <- readRDS(file.path(ROOT_OUT_FP, "taxa.rds"))
  
  r1.map <- qiimer::read_qiime_mapping_file(
    file.path(ROOT_DATA_FP, "MIRA1/20171110_kelly_mira_mapping_file.txt")) %>%
    select(specimen_id=SampleID, specimen_type=SampleType, flow_cell_id) %>%
    mutate(specimen_id2 = paste0(specimen_id, "_M1"))
  r2.map <- qiimer::read_qiime_mapping_file(
    file.path(ROOT_DATA_FP, "MIRA2/kelly_mira2_mapping_file.tsv")) %>%
    select(specimen_id=SampleID, specimen_type=SampleType, flow_cell_id) %>%
    mutate(specimen_id2 = paste0(specimen_id, "_M2"))
  r3.map <- qiimer::read_qiime_mapping_file(
    file.path(ROOT_DATA_FP, "MIRA3/20180228_kelly_mira3_mapping_file.txt")) %>%
    select(specimen_id=SampleID, specimen_type=SampleType, flow_cell_id) %>%
    mutate(specimen_id2 = paste0(specimen_id, "_M3"))
  mapped.samples <- bind_rows(r1.map, r2.map, r3.map) %>%
    mutate(specimen_id = str_replace_all(specimen_id, "\\.", "_"))
  
  labvantage <- readxl::read_excel(file.path(ROOT_DATA_FP, "mira_labvantage.xls")) %>%
    select(
      subject_id=`Ext Participant ID`, 
      specimen_id=`Immediate Parent Sample ID`, 
      orig_specimen_type=`Original Sample Type`,
      collection_date=`Collection Date`) %>%
    mutate(specimen_id = stringr::str_replace_all(specimen_id, "-", "_"))
  
  samples <- left_join(mapped.samples, labvantage) %>%
    group_by(subject_id) %>%
    mutate(study_day = as.integer(difftime(collection_date, min(collection_date), units="days")))
  
  otus <- data.frame(seq=colnames(seqtab)) %>% mutate(otu_id = seq_len(ncol(seqtab)))
  colnames(seqtab) <- otus$otu_id
  agg <- reshape2::melt(seqtab, varnames=c("specimen_id2", "otu_id"), value.name="read_count") %>%
    left_join(samples)
  
  list(
    otus=otus,
    samples,
    taxa=taxa,
    agg=agg
  )
}

mira <- load_all_data()

```

## Summary plots

### Pileups by subjects

The following plots show the number of ASVs common to any number of subjects. The first shows 
all counts of ASVs in any number of subjects, while the second shows the number of ASVs common to
ten or more subjects.

```{r}
invisible(within(mira, {
  asvs_by_subject <- agg %>% 
    select(otu_id, subject_id, study_day, specimen_type=orig_specimen_type, read_count) %>%
    filter(read_count > 0) %>%
    filter(specimen_type %in% c("Oral Swab", "Stool Swab", "Sputum")) %>%
    group_by(otu_id, specimen_type) %>%
    summarize(distinct_subjects = n_distinct(subject_id)) %>%
    group_by(specimen_type, distinct_subjects) %>% 
    summarize(distinct_otus = n_distinct(otu_id))
  
  p.asv_pileup_all <- ggplot(asvs_by_subject, aes(distinct_subjects, distinct_otus)) + geom_bar(stat="identity") +
    facet_grid(~ specimen_type, scales="free_x", space="free") +
    scale_y_continuous(limits=c(0, 8e3), expand=c(0,0)) +
    labs(x="Unique subjects", y="Unique ASVs", title="Counts of ASVs in any number of subjects")
  ggsave(p.asv_pileup_all, filename = file.path(FIG_FP, "ASV_Pileup_All.pdf"), width=5, height=3)
  plot(p.asv_pileup_all)
  p.asv_pileup_gt10 <- ggplot(
    filter(asvs_by_subject, distinct_subjects > 10), 
    aes(distinct_subjects, distinct_otus)) + geom_bar(stat="identity") +
    facet_grid(. ~ specimen_type, scales="free_x", space="free") +
    scale_y_continuous(limits=c(0,60), expand=c(0,0)) +
    labs(x="Unique subjects", y="Unique ASVs", title="Counts of ASVs in more than 10 subjects")
  ggsave(p.asv_pileup_gt10, filename = file.path(FIG_FP, "ASV_Pileup_10_Subjects.pdf"), width=5, height=3)
  plot(p.asv_pileup_gt10)
}))
```

### Pileups by earliest detection

The next two plots show the number of ASVs according to their first detection date. The first plot
includes day 0 (i.e. ASVs endogenous to the patient) while the next plot shows all ASVs that appear
only after day 0 (i.e. potentially hospital-acquired).

```{r}
invisible(with(mira, {
  asvs_by_day <- agg %>% 
    select(otu_id, subject_id, study_day, specimen_type=orig_specimen_type, read_count) %>%
    filter(read_count > 0) %>%
    filter(specimen_type %in% c("Oral Swab", "Stool Swab", "Sputum")) %>%
    group_by(otu_id, specimen_type) %>%
    summarize(earliest_study_day = min(study_day)) %>%
    mutate(earliest_study_day = ifelse(earliest_study_day > 200, 200, earliest_study_day)) %>%
    group_by(earliest_study_day, specimen_type) %>%
    summarize(count = n())
  p.all_days <- ggplot(asvs_by_day, aes(earliest_study_day, count)) + 
    geom_bar(stat="identity", fill="black") +
    facet_grid(. ~ specimen_type, space="free", scales="free") +
    scale_y_continuous(expand=c(0,0)) +
    scale_x_continuous(limits=c(-1, 201), breaks = seq(0, 200, by=50), labels=c(0, 50, 100, 150, "200+")) + 
    labs(y="Unique ASVs", x="Earliest study day", title="ASVs by earliest detection")
  ggsave(p.all_days, file=file.path(FIG_FP, "ASV_Pileup_Earliest_Day.pdf"), width=5, height=3)
  plot(p.all_days)
  p.all_except_0 <- ggplot(filter(asvs_by_day, earliest_study_day > 0), aes(earliest_study_day, count)) + 
    geom_bar(stat="identity") +
    scale_y_continuous(expand=c(0,0)) +
    scale_x_continuous(limits=c(0, 201), breaks = c(1, seq(50, 200, by=50)), labels=c(1,50,100,150, "200+")) + 
    facet_grid(. ~ specimen_type, space="free", scales="free") +
    labs(y="Unique ASVs", x="Earliest study day", title="ASVs by earliest detection (excluding day 0)")
  ggsave(p.all_except_0, file=file.path(FIG_FP, "ASV_Pileup_Earliest_Day_Except_D0.pdf"), width=5, height=3)
  plot(p.all_except_0)
}))
```


